---
title: "Unveiling the signaling network of FLT3-ITD AML improves drug sensitivity prediction"
subtitle: "FLT3-ITD patients: RNAseq analysis of 262 genes"
author: "Veronica Venafra"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
source('0.libraries.R')
```

## From BAM to raw counts
Gene counts tables were generated from bam files using Rsubread. 
```{r eval = FALSE}

# Path of hard disk where company data were stored
hd_path <- '/Volumes/EXT_DISK/AML_FLT3_Sequencing/BWA_mapping/'
patients <- c('AR', 'DN', 'DNF', 'FE', 'FF', 'LS', 'MD', 'NG', 'RL_1',
              'RL_2', 'SE', 'SF', 'SM', 'TR')

fc <- featureCounts(files = paste0(hd_path, patients, '.deduped.bam'),
              annot.inbuilt = 'hg38', 
              isPairedEnd = TRUE, nthreads = 12)

df <- fc$counts
df <- data.frame(df) %>% rownames_to_column() %>% as_tibble() %>% 
  dplyr::rename(GeneID = rowname)

# keep only genes with at least one read
genes <- df %>% filter(!if_all(.cols = where(is.numeric), ~ .x == 0))

# translate ENSEMBL ID in Gene name
gene_name <- getSYMBOL(x = genes$GeneID, data = 'org.Hs.eg.db')
mapping_df <- tibble(GeneID = names(gene_name), gene_name)
genes <- left_join(genes, mapping_df, by = 'GeneID') %>% 
  relocate(GeneID, gene_name) %>%
  arrange(gene_name)

write_tsv(genes, './result/rnaseq/raw_count_matrix.tsv')
```

## Annotate sample according to ITD location

```{r eval = FALSE}

# 1) countData derived from Rsubread package (BAM files -> to raw counts)
cts <- read_tsv('./result/rnaseq/raw_count_matrix.tsv')
classified <- read_tsv('./result/getitd/getitd_compact_res_custom.tsv')

classified$type[grepl('JMD\\+', classified$mutation)] <- paste0('JMD_TKD', 1:sum(grepl('JMD\\+', classified$mutation)))
classified$type[grepl('JMD$', classified$mutation)] <- paste0('JMD', 1:sum(grepl('JMD$', classified$mutation)))
classified$type[grepl('^TKD$', classified$mutation)] <- paste0('TKD', 1:sum(grepl('^TKD$', classified$mutation)))

write_tsv(classified, './result/getitd/mapping_getitd_patientID.tsv')

colnames(cts)[3:length(colnames(cts))] <- str_sub(colnames(cts)[3:length(colnames(cts))], 1,4)
colnames(cts)[3:length(colnames(cts))] <- str_remove_all(colnames(cts)[3:length(colnames(cts))], '.d')
colnames(cts)[3:length(colnames(cts))] <- str_remove_all(colnames(cts)[3:length(colnames(cts))], '\\.')
colnames(cts)[3:length(colnames(cts))] <- str_replace_all(colnames(cts)[3:length(colnames(cts))], '_', '.')

#cts <- cts %>% select(-c('RL.2'))

#rename columns according to ITD classification
v <- classified$type[match(colnames(cts)[3:length(colnames(cts))], classified$sample)]
colnames(cts)[3:length(colnames(cts))] <- v
write_tsv(cts, './result/rnaseq/count_matrix_flt3_annotated.tsv')

# 2) colData construction
colData <- data.frame(condition = rep('a', length(v)))
rownames(colData) <- v

colData$condition[grepl('^JMD', rownames(colData))] <- 'JMD'
colData$condition[grepl('^TKD', rownames(colData))] <- 'TKD'
colData$condition[grepl('^JMD_', rownames(colData))] <- 'JMD_TKD'

# create cts1 with columns same as colData
cts1 <- cts %>% column_to_rownames('GeneID') 
cts1 <- cts1[, rownames(colData)]

all(rownames(colData) == colnames(cts1))

# create deseqdataset
dds <- DESeqDataSetFromMatrix(countData = cts1,
                              colData = colData,
                              design = ~ condition)

# add as features the gene_names of geneID
featureData <- data.frame(gene = cts$gene_name)
rownames(featureData) <- rownames(cts1)

mcols(dds) <- DataFrame(mcols(dds), featureData)

write_rds(dds, './result/rnaseq/dds_object.rds')
```

## Normalization with edgeR and z-score computation

```{r eval = FALSE}

# Read input data
cts <- read_tsv('./result/deseq2/count_matrix_flt3_annotated.tsv')
dds <- readRDS('./result/deseq2/dds_object.rds')

cts$gene_name <- NULL
cts <- cts %>% 
  column_to_rownames('GeneID')

# Counts threshold to obtain exactly the number of sequenced genes 262
keep <- rowSums(counts(dds) >= 3950) >= 8
dds1 <- dds[keep,]

raw_counts <- counts(dds1, normalized=FALSE)

# Define groups
group <- dds$condition

y <- DGEList(counts=raw_counts,group=group)
y <- calcNormFactors(y)

# Get normalized counts
norm_counts <- data.frame(y$counts)

# Logarithmic transformation
norm_counts_log2 <- data.frame(log2(norm_counts)) %>% 
  rownames_to_column('geneID')

gene_IDs <- tibble(geneID = rownames(mcols(dds1)), 
                   gene_name = mcols(dds1)$gene)

norm_counts_log2 <- inner_join(norm_counts_log2, 
                               gene_IDs) %>% 
  relocate(geneID, gene_name)

# Compute z-score
mean_v <- colMeans(norm_counts_log2[3:ncol(norm_counts_log2)])
sd_v <- colSds(as.matrix(norm_counts_log2[sapply(norm_counts_log2, is.numeric)]))
names(sd_v) <- names(mean_v)

norm_counts_log2_z <- norm_counts_log2
for(patient in names(mean_v)){
  norm_counts_log2_z[, patient] <- (norm_counts_log2[,patient] - mean_v[patient])/sd_v[patient] 
}

write_tsv(norm_counts_log2_z, './result/rnaseq/zscore.tsv')
```

## Visualization

### Clustering 

```{r, fig.height=3, fig.width=5, fig.align='center'}

norm_counts_log2_z <- read_tsv('./result/rnaseq/zscore.tsv') 
norm_counts_log2_z <- norm_counts_log2_z %>% dplyr::select(-c('geneID'))

norm_counts_log2_z <- norm_counts_log2_z %>% column_to_rownames('gene_name')
norm_counts_log2_z_t <- t(norm_counts_log2_z)

z_dist = dist(norm_counts_log2_z_t, method = 'euclidean')
z_hc = hclust(z_dist, method = 'complete')

z_plot <- ggdendrogram(z_hc, theme_dendro = TRUE) +
  theme(axis.line.y = element_line(colour = 'darkgrey'),
        axis.ticks.y = element_line()) 
z_plot
```

**Figure 5D** Hierarchical clustering of patients according to their expression profile of 262 genes.

### PCA

```{r eval = FALSE}

# Read DeSeq2 object
dds <- readRDS('./result/rnaseq/dds_object.rds')

# threshold to obtain exactly the number of sequenced genes 262
keep <- rowSums(counts(dds) >= 3950) >= 8
dds1 <- dds[keep,]
rld <- rlog(dds, blind = TRUE)

p <- plotPCA1(rld)

PCA <- p +
  theme_classic() +
  scale_color_brewer(palette = 'Dark2') +
  theme(legend.title = element_blank())

write_rds(PCA, './result/rnaseq/PCA.rds')
```

```{r, fig.height=4, fig.width=5, fig.align='center'}
PCA_plot <- read_rds('./result/rnaseq/PCA.rds')
PCA_plot
```

**Figure S6C** Principal Component Analysis (PCA) of patients’ expression profile.


