---
title: "Unveiling the signaling network of FLT3-ITD AML improves drug sensitivity
  prediction"
subtitle: "Network visualization" 
author: "Veronica Venafra"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source('0.libraries.R')
```

The optimized models were visualized on Cytoscape. The main steps were:

1.  Create the files in R.

2.  Load the files in Cytoscape.

## 1. File generation in R

To load the models in Cytoscape you need:

-   a *sif file* representing the structure of the optimized model
-   a file of the *nodes attributes* specifying which species is a
    sentinel, an inhibited/stimulated protein
-   a file of the *edge attributes* specifying the frequency of each
    edge in the family of optimized model.

### Generation of sif file and edges attributes

To generate the sif file of the network and the edge attribute file we
used the **writeScaffold** function of CNOptR package, rewritten to
avoid some errors. The code is available in `0.libraries.R` file.

The network file will be: \*[cellline]\_Scaffold.sif *The edge attribute
file will be:* [cellline]\_weightsScaffold.EA\*

```{r eval = FALSE}

for(cellline in c('JMD', 'TKD')){
  
  # Load the family of 100 best model to compute the edge frequency
  load(paste0('./results/', cellline, '_subfamily_threshold.RData'))
  
  # Assign to the train object in the RData environment the family of 100 models
  # stored in optModels variable. In the writeScaffold function the frequency
  # of each logic gate in the family of models will be computed.
  train$stringsTol <- data.matrix(optModels)
  
  writeScaffold(
  modelOriginal = pknmodel,
  modelComprExpanded = modelT0,
  optimResT1 = train,
  optimResT2 = NA,
  CNOlist = CNOlist,
  tag = paste0('./results/visualization/', cellline))
}
```

### Generation of nodes attributes

The nodes attribute file will be: \*[cellline]\_nodesPKN.attrs\* To
generate that you have to run the code below and then modify the
extension of the generate file in '.attrs' and modify the header in
'Cytocopter.NodesType'.

```{r eval = FALSE}
pkn_path <- './input/PKN_3008.sif'
pknmodel <- readSIF(pkn_path)

# get normalized experimental data
CNOlist <- CNOlist(paste0('./input/MD_', cellline, '_NORM.csv'))

# compress and expand the PKN
modelT0 <- preprocessing(CNOlist, pknmodel, expansion = TRUE)

# execute the function only to obtain the nodes attributes file!
writeNetwork(
  modelOriginal = pknmodel,
  modelComprExpanded = modelT0,
  optimResT1 = train,
  CNOlist = CNOlist,
  tag = paste0('./results/visualization/', cellline))
```

## 2. Load files in Cytoscape

Once obtained all the files, we:

1.  Loaded the ***[cellline]\_scaffold.sif.***

2.  Imported as nodes attributes the ***[cellline]\_nodesPKN.attrs***

3.  Imported as edge attributes the
    ***[cellline]\_weightsScaffold.attrs***

4.  The Cytocopter style was used to color nodes and edges

5.  To obtain the high confidence models we filtered in Cytoscape the
    edges with Weight \> 0.4.
