---
title: "Unveiling the signaling network of FLT3-ITD AML improves drug sensitivity
  prediction"
subtitle: "Combinatory treatments inference"
author: "Veronica Venafra"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source('0.libraries.R')
```

To identify promising co-treatments able to revert the resistant
phenotype, we exploited the predictive power of the generated Boolean
models and performed an *in silico* knock-out of key kinases present in
the FLT3 ITD-TKD model (ERK1/2, MEK1/2, GSK3A/B, IGF1R, JNK, KRAS,
MEK1/2, MTOR, PDPK1, PI3K, p38)

### Steady state computation upon each combinatorial inhibition condition

```{r, eval = FALSE}

for(cellline in c('JMD', 'TKD')){
  
  # Load the family of 100 best models
  load(paste0('./results/', cellline, '_subfamily_threshold.RData'))
  
  # Get a vector of inhibited kinases in experimental design
  inhibited <- colnames(CNOlist@inhibitors)
  
  # Define the simulation conditions that are inhibited kinases in the 
  # experimental design and four more key signaling kinases
  conditions <- c('tumor', inhibited,
                  'ERK1_2', 'KRAS',
                  'PDPK1', 'IGF1R')
  
  for (condition in conditions){
    
    # Define as a baseline that the starting point are all set to 1
    CNOlist@cues[1,1:3] <- rep(1, 3)
    CNOlist@cues <- t(as.matrix(CNOlist@cues[1,1:3]))
    
    d <- c(1,1,1)
    names(d) <- c('FLT3', 'TNF', 'IGF1R')
    CNOlist@stimuli <-  t(as.matrix(d))
    
    if(condition == 'tumor'){
      # -- TUMOR SIMULATION -- #
      # Remove the FLT3 inhibitor from the design
      CNOlist@inhibitors[1,1] <- 0
      CNOlist@inhibitors <- t(as.matrix(CNOlist@inhibitors[1,1]))
    }else if(condition == 'FLT3'){
      # -- FLT3 INHBITION CONDITION -- #
      # Add the inhibition of FLT3 in inhibitors vector
      in_val <- c(1)
      names(in_val) <- condition
      CNOlist@inhibitors <- t(as.matrix(in_val))
    }else{
      ## -- FLT3 INHIBITION + ONE TARGET KINASE -- ##
      # Add the inhibition of FLT3 and a kinase defined in design vector
      in_val <- c(1, 1)
      names(in_val) <- c('FLT3', condition)
      CNOlist@inhibitors <- t(as.matrix(in_val))
    }
    
    ### -- MANUAL SIMULATION -- ##
    
    # creating object for simulation
    simList <- prep4sim(modelT0) 
    
    matrix_sim_q <- list()
    for(i in c(1)){ #select only the best model
     
      bString <- as.vector(unlist(best_100_models[i,]))
    
      # cutting expanded gates to optimal bString
      modelCut <- cutModel(modelT0, list(bString)[[1]])
      
      # creating a new simList from cutted model
      newSimList <- cutSimList(simList, list(bString)[[1]])
      
      # creating index list
      indexList <- indexFinder(CNOlist, modelT0)
      
      ## ** SIMULATION A T0 (before treatment) ** 
      simResT0 <- simulatorT1(CNOlist=CNOlist, model=modelCut, simList=newSimList,
                              indexList=indexList, mode = 0)
      colnames(simResT0) <- modelT0$namesSpecies
      
      ## ** SIMULATION A T1 (after treatment) **
      simResT1 <- simulatorT1(CNOlist=CNOlist, model=modelCut, simList=newSimList, 
                              indexList=indexList, mode = 1)
      colnames(simResT1) <- modelT0$namesSpecies
      
      ### -- STORING RESULTS in data objects -- ###
      # list object
      simRes <- list(T0 = simResT0, T1 = simResT1)
      
      if(sum(is.na(simRes$T0)) > 0 | sum(is.na(simRes$T1)) > 0){
        matrix_sim_q[[i]] <- list()
      }else{
        matrix_sim_q[[i]] <- list()
        matrix_sim_q[[i]] <- simRes
      }
    }
    
    matrix_sim_t0 <- lapply(matrix_sim_q, function(x){x$T0})
    matrix_sim_t0 <- Filter(Negate(is.null), matrix_sim_t0)
    averaget0 <-  Reduce("+", matrix_sim_t0) / length(matrix_sim_t0)
    
    matrix_sim_t1 <- lapply(matrix_sim_q, function(x){x$T1})
    matrix_sim_t1 <- Filter(Negate(is.null), matrix_sim_t1)
    averaget1 <-  Reduce("+", matrix_sim_t1) / length(matrix_sim_t1)
    
    # matrix object
    sim_matrix <- matrix(nrow = length(simRes$T0), ncol = 2)
    
    sim_matrix[,1] <- t(as.matrix(averaget0)) 
    sim_matrix[,2] <- t(as.matrix(averaget1))
    rownames(sim_matrix) <- colnames(simResT1)
    colnames(sim_matrix) <- c('T0', paste0('FLT3i+',condition,'i'))
    
    if(condition == 'tumor'){
      sim_matrix_new <- sim_matrix
      conditions_names <- c('T0', 'FLT3-ITD+TNF+IGF')
      colnames(sim_matrix_new) <- conditions_names
    }else{
      conditions_names <- colnames(sim_matrix_new)
      sim_matrix_new <- cbind(sim_matrix_new, sim_matrix[,2])
      if(condition == 'FLT3'){
        colnames(sim_matrix_new) <- c(colnames(sim_matrix_new)[1:length(colnames(sim_matrix_new))-1], 
                                      paste0(condition,'i'))
      }else{
        colnames(sim_matrix_new) <- c(colnames(sim_matrix_new)[1:length(colnames(sim_matrix_new))-1], 
                                      paste0('FLT3i+',condition,'i'))
      }
    }
  }
  
  sim_matrix_new <- sim_matrix_new[,-1] 
  
  if(cellline == 'TKD'){
    sim_matrix_TKD <- sim_matrix_new
  }else{
    sim_matrix_JMD <- sim_matrix_new
  }
  
}

sim_tibble_JMD <- tibble(rownames_to_column(data.frame(sim_matrix_JMD), 'node'))
write_tsv(sim_tibble_JMD, './results/JMD_combinations_simulation_steady_state.tsv')

sim_tibble_TKD <- tibble(rownames_to_column(data.frame(sim_matrix_TKD), 'node'))
write_tsv(sim_tibble_TKD, './results/TKD_combinations_simulation_steady_state.tsv')
```

### Inference of Proliferation and Apoptosis

Then we inferred the activity of **APOPTOSIS** and **PROLIFERATION**
phenotypes. We eventually selected co-treatments in the FLT3 ITD-TKD
model able to trigger activation levels of the 'apoptosis' and
'proliferation' to the same level as the FLT3 ITD-JMD model.

```{r, eval = FALSE}

sim_tibble_JMD <- read_tsv('./results/JMD_combinations_simulation_steady_state.tsv')
sim_tibble_TKD <- read_tsv('./results/TKD_combinations_simulation_steady_state.tsv')

phens <- c('APOPTOSIS', 'PROLIFERATION')
plot_list <- list()

for(phen_i in 1:length(phens)){
  phen <- phens[phen_i]

  apoptosis_annotation <- read_tsv('./results/phenotypes_analysis/paths_to_apoptosis_proliferation.txt') %>%
    select(node = QueryNode, phenotype = EndNode, apoptosis = Final_Effect)
  
  apoptosis_annotation <- apoptosis_annotation %>% 
    filter(phenotype == phen) %>% 
    select(node, apoptosis)  %>%
    distinct()
  
  for(i in c(1:(ncol(sim_tibble_JMD)-3))){
    
    j = i + 3
    conditions_sim <- colnames(sim_tibble_JMD)[j]

    for(cellline in c('JMD', 'TKD')){
      
      if(cellline == 'JMD'){
        annotated <- left_join(sim_tibble_JMD, apoptosis_annotation %>% 
                                 select(node, apoptosis), by = 'node')
        
        if(phen == 'APOPTOSIS'){
          annotated <- annotated %>% 
            filter(node %in% c('MTOR', 'STAT5A'))
        }else{
          annotated <- annotated %>% 
            filter(node %in% c('PI3K', 'MAPK14', 'STAT5A', 'MTOR'))
          
        }
        
        
      }else{
        annotated <- left_join(sim_tibble_TKD, apoptosis_annotation %>% 
                                 select(node, apoptosis), by = 'node')
        
        if(phen == 'APOPTOSIS'){
          annotated <- annotated %>%
            filter(node %in% c('STAT5A', 'AKT'))
        }else{
          annotated <- annotated %>%
            filter(node %in% c('PI3K', 'KRAS', 'CREB1', 'STAT5A'))
        }
      }
      
      annotated <- annotated %>% 
        select('node', T0 = 2, T10 = 3, T90 = j, apoptosis) %>%
        filter(!is.na(apoptosis))
      
      apo <- annotated %>% 
        filter(!is.na(apoptosis)) %>%
        mutate(T0_apo = T0*apoptosis, 
               T10_apo = T10*apoptosis,
               T90_apo = T90*apoptosis) %>% 
        group_by(apoptosis) %>% 
        summarise(T0 = sum(T0_apo),
                  T10 = sum(T10_apo),
                  T90 = sum(T90_apo))
      
        apo <- apo %>% summarise(T0 = sum(T0), 
                                 T10 = sum(T10),
                                 T90 = sum(T90))
        
      if(cellline == 'JMD'){
        apo$cellline <- 'JMD'
        apo_JMD <- apo
        annotated_JMD <- annotated
      }else{
        apo$cellline <- 'TKD'
        apo_TKD <- apo
        annotated_TKD <- annotated
      }
    }
    apo_tb <- bind_rows(apo_JMD, apo_TKD)
    
    if(i == 1){
      apo_tb_f <- apo_tb
    }else{
      sub_apo_tb <- apo_tb %>% select(T90, cellline)
      colnames(sub_apo_tb) <- c(conditions_sim, 'cellline')
      apo_tb_f <- left_join(apo_tb_f, sub_apo_tb, by = 'cellline')
    }
  }
  
  apo_tb_f <- apo_tb_f %>% relocate(cellline)

  apo_matrix <- as.matrix(apo_tb_f %>% 
                            column_to_rownames('cellline'))
  
  a <- data.frame(t(apo_matrix)) %>% rownames_to_column()
  JMD <- a[, c('rowname', 'JMD')]
  JMD$rowname <- c('FLT3.TNF.IGF1R', 'FLT3i', 'FLT3i.MAPK14i', JMD$rowname[4:length(rownames(JMD))])
  colnames(JMD) <- c('time', 'apoptosis')
  JMD$cellline <- 'JMD'
  
  TKD <- a[, c('rowname', 'TKD')]
  TKD$rowname <- c('FLT3.TNF.IGF1R', 'FLT3i', 'FLT3i.MAPK14i', TKD$rowname[4:length(rownames(TKD))])
  colnames(TKD) <- c('time', 'apoptosis')
  TKD$cellline <- 'TKD'
  
  barplot_tb <- bind_rows(JMD, TKD) %>% 
    mutate_at('apoptosis', as.numeric)
  
  if(phen == 'APOPTOSIS'){
    barplot_tb$apoptosis <- barplot_tb$apoptosis * -1
  }
  
  barplot_tb_sub <- barplot_tb
  
  barplot_apoptosis <- ggplot(barplot_tb_sub, aes(x = time, y = apoptosis, fill = cellline)) +
    geom_bar(stat = 'identity', position = 'dodge', width = 0.8) +
    xlab('') + ylab(phen)+
    scale_fill_manual(values = c('#36437F','#E1BD6A'))+
    theme_classic() +
    ylim(0,7)+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 10),
          axis.text.y = element_text(size = 10), 
          axis.title = element_text(size = 12))
  
  plot_list[[phen_i]] <- barplot_apoptosis
}

write_rds(plot_list, './results/plot_list_combinations.rds')
```

```{r, fig.height=5, fig.width=8, fig.align='center'}

plot_list <- readRDS('./results/plot_list_combinations.rds')
n <- length(plot_list)
nCol <- floor(sqrt(n))
p <- do.call("grid.arrange", c(plot_list, ncol=nCol))

```

**Figure 4B-C.** Bar plot showing the *in silico* simulation of
proliferation activation and apoptosis inhibition levels in untreated
and FLT3i conditions in combination with knock-out of each of 10 crucial
kinases in FLT3 ITD-JMD (blue) and -TKD (yellow) cells. 
